The Proofs

### 2024-07-16 proofs/content.tex
WIP: fix references, add appendix

\newcommand{\proofGenDir}{_gen}

\section{Declarative Type Systems}

\subsection{Grammar}
We assume that there is an infinite set of positive and 
negative \emph{type} variables. Positive type variables are denoted as 
$[[α⁺]]$, $[[β⁺]]$, $[[γ⁺]]$, etc.
Negative type variables are denoted as $[[α⁻]]$, $[[β⁻]]$, $[[γ⁻]]$, etc.
We assume there is an infinite set of \emph{term} variables,
which are denoted as $[[x]]$, $[[y]]$, $[[z]]$, etc.
A list of objects (variables, types or terms) is denoted by
an overline arrow. For instance, $[[pas]]$ is a list of positive type variables, 
$[[nbs]]$ is a list of negative type variables, 
$[[args]]$ is a list of values, which are arguments of a function.
$[[fv(iP)]]$ and $[[fv(iN)]]$ denote the set of free variables 
in a type $[[iP]]$ and $[[iN]]$, respectively.

\bigskip

% \ottgrammartabular{
%   \ottiP\ottinterrule
%   \ottiN\ottinterrule
% }

\begin{definition}[Declarative Types]
  \hfill
  \begin{multicols}{2}
    \ottgrammartabular{
      \ottiN\ottinterrule
    }

    \ottgrammartabular{
      \ottiP\ottinterrule
    }
    \columnbreak
  \end{multicols}
\end{definition}

\subsection{Equalities}
For simplicity, we assume alpha-equivalent terms are equal.
This way, we assume that substitutions do not capture bound variables.
Besides, we equate
$[[∀pas.∀pbs.iN]]$ with $[[∀pas,pbs.iN]]$, 
as well as $[[∃nas.∃nbs.iP]]$ with $[[∃nas,nbs.iP]]$,
and lift these equations transitively and congruently 
to the whole system.

\subsection{Contexts and Well-formedness}

@@

\subsection{Substitutions}

@@

\subsection{Declarative Subtyping}
Subtyping is one of the key mechanisms of our system. 
It realizes the polymorphism: abstract $[[∀]]$ and 
$[[∃]]$ types can be used where concrete types are expected,
exactly because of the subtyping relation between them.

@@

\section{Algorithmic Type System}

\subsection{Grammar}

@@

\subsection{Fresh Variable Selection}
\label{sec:fresh-selection}
Both the subtyping and the type inference algorithm
rely on the ability to select fresh, unused variables.
For a set of variables $[[varset]]$, it is indicated as 
$[[varset are fresh]]$ in the inference rules.
We assume that the selection subroutine always succeeds and is 
deterministic. In other words, whenever it is called in 
an algorithmic inference rule, it returns the same result, 
uniquely determined by the input of this rule.

\subsection{Variable Algorithmization}
\label{sec:variable-algorithmization}

@@

\subsection{Contexts and Well-formedness}

@@

\subsection{Subsitutions}

@@

\subsection{Equivalence and Normalization}
\label{sec:equivalence-normalization}

@@

The rest of this chapter is devoted to the
central algorithm of the type system---the subtyping algorithm. 
\Cref{fig:subtyping-algo} shows the dependency graph of the subtyping algorithm.
The nodes represent the algorithmic procedures, and the edge $A \to B$ means that 
$A$ uses $B$ as a sub-procedure.

\begin{figure}[t]
  \centering
  \begin{tikzpicture}
    [>={Stealth[scale=2]},node distance=2.4cm,every node/.style={draw,rectangle},every text node part/.style={align=center}]

    % Define nodes
    \node[] (1) {Negative Subtyping\\$[[Γ ; Θ ⊨ uN ≤ iM ⫤ SC]]$\\(\cref{sec:subtyping})};
    \node[below of=1] (3) {Positive Subtyping\\$[[Γ ; Θ ⊨ uP ≥ iQ ⫤ SC]]$\\(\cref{sec:subtyping})};
    \node[left=1cm of 3] (2) {Subtyping Constraint Merge\\$[[Θ ⊢ SC1 & SC2 = SC3]]$\\(\cref{sec:constraint-merge})};
    \node[right=1.4cm of 3] (5) {Unification\\ $[[Γ ; Θ ⊨ uN ≈u iM ⫤ UC]]$\\ $[[Γ ; Θ ⊨ uP ≈u iQ ⫤ UC]]$\\(\cref{sec:unification})};
    \node[below of=3] (4) {Upgrade\\$[[upgrade Γ ⊢ iP to Δ = iQ]]$\\(\cref{sec:lub})};
    \node[below of=2] (6) {Least Upper Bound\\$[[Γ ⊨ iP1 ∨ iP2 = iQ]]$\\(\cref{sec:lub})};
    \node[below of=6] (7) {Anti-Unification\\$[[Γ ⊨ iP1 ≈au iP2 ⫤ ( Ξ , uQ , aus1 , aus2 )]]$\\$[[Γ ⊨ iN1 ≈au iN2 ⫤ ( Ξ , uM , aus1 , aus2 )]]$\\(\cref{sec:antiunification})};
    \node[below of=5] (8) {Unification Constraint Merge\\$[[Θ ⊢ UC1 & UC2 = UC3]]$\\(\cref{sec:constraint-merge})};

    % Define edges
    \draw[->] (1) to (2);
    \draw[->] (1) to (3);
    \draw[->] (1) to (5);
    \draw[->] (2) to (3);
    \draw[->] (2) to (6);
    \draw[->] (3) to (4);
    \draw[->] (3) to (5);
    \draw[->] (4) to (6);
    \draw[->] (5) to (8);
    \draw[->] (6) to (7);
  \end{tikzpicture}  
  \caption{Dependency graph of the subtyping algorithm}
  \label{fig:subtyping-algo}
\end{figure}

\subsection{Subtyping}
\label{sec:subtyping}

@@

Let us overview the inference rules of the subtyping algorithm.
\begin{itemize}
  \item \ruleref{\ottdruleANVarLabel} and \ruleref{\ottdruleAPVarLabel} 
    are the base cases. They copy the corresponding declarative rules and
    ensure reflexivity.
  \item \ruleref{\ottdruleAPUVarLabel} is the only case generating 
    subtyping constraints. In this case, we must ensure
    that the resulting constraints guarantee that the instantiation of 
    $[[â⁺]]$ is a supertype of $[[iP]]$.
    However, the obvious constraint $[[â⁺ :≥ iP]]$ might be problematic
    if  $[[iP]]$ is not well-formed in $[[Θ(â⁺)]]$. For this reason,
    we use the \emph{upgrade} procedure (it will be covered in \cref{sec:lub})
    to find the minimal supertype of $[[iP]]$, which is well-formed in $[[Θ(â⁺)]]$.

@@

  \item \ruleref{\ottdruleAShiftDLabel} and \ruleref{\ottdruleAShiftULabel} are the
    \emph{shift} rules. According to the declarative system,
    shifted subtyping requires equivalence. In the presence of the algorithmic 
    variables, it means that the left and the right-hand sides of the subtyping
    must be unified. Hence, the shift rules invoke the unification algorithm, 
    which will be discussed in \cref{sec:unification}. The unification 
    returns the minimal set of constraints $[[UC]]$, which is necessary
    and sufficient for the subtyping.

  \item \ruleref{\ottdruleAArrowLabel}.
    In this case, the algorithm makes two calls:
    a recursive call to the negative subtyping algorithm for the argument types,
    and a call to the positive subtyping algorithm for the result types.
    After that, the resulting constraints are merged using the
    \emph{subtyping constraint merge} procedure, 
    which is discussed in \cref{sec:constraint-merge}.
  \item \ruleref{\ottdruleAForallLabel} and \ruleref{\ottdruleAExistsLabel}
    are symmetric. These are the only places where 
    the algorithmic variables are introduced.
    It is done by algorithmization (\cref{sec:variable-algorithmization}) 
    of the quantified variables: these variables are replaced by 
    fresh algorithmic variables in the body of the quantified type,
    the algorithmic variables are added to the subtyping context $[[Θ]]$,
    after that, the recursive call is made. Notice that the declarative context
    $[[Γ]]$ is extended by the quantified variables from the right-hand side,
    which matches the declarative system.
\end{itemize}

@@

\subsection{Constraints}

@@

\subsubsection{Constraint Satisfaction}

@@

\subsubsection{Constraint Merge}
\label{sec:constraint-merge}

@@

  \item \ruleref{\ottdruleSCMESupSupLabel} 
    In this case, we find the least upper bound of the types from the input
    restrictions, 
    and as the output, restrict the variable to be a supertype of the result.
    The least upper bound procedure will be discussed in \cref{sec:lub}.
\end{itemize}

@@

\subsection{Unification}
\label{sec:unification}

@@

\begin{itemize}
  \item \ruleref{\ottdruleUShiftULabel}, \ruleref{\ottdruleUShiftDLabel}, 
    \ruleref{\ottdruleUForallLabel}, and \ruleref{\ottdruleUExistsLabel}
    are defined congruently. In the shift rules, the algorithm
    removes the outermost constructor. In the
    $[[∀]]$ and $[[∃]]$ rules, it removes the quantifiers,
    adding the quantified variables to the context $[[Γ]]$.
    Notice that $[[Θ]]$, which specifies
    the contexts in which the algorithmic variables must be instantiated,
    is not changed.
  \item \ruleref{\ottdruleUNVarLabel} and \ruleref{\ottdruleUPVarLabel} 
    are the base cases. 
    Since the sides are equal and free from algorithmic variables,
    the unification returns an empty constraint. 
  \item \ruleref{\ottdruleUNVarLabel} and \ruleref{\ottdruleUPVarLabel}
    are symmetric cases constructing the constraints. 
    When an algorithmic variable is unified with a type, 
    we must check that the type is well-formed in the required context,
    and if it is, we return a constraint restricting the variable
    to be equivalent to that type.
  \item \ruleref{\ottdruleUArrowLabel}.
    In this case, the algorithm makes two recursive calls:
    it unifies the arguments and the results of the arrows.
    After that, the resulting constraints are merged using the
    \emph{unification constraint merge} procedure, 
    which is discussed in \cref{sec:constraint-merge}.
    Notice that $[[UC1]]$ and $[[UC2]]$ are guaranteed to be
    \emph{unification} constraints, not arbitrary \emph{subtyping} 
    constraints: it is important for modularizing the proofs, 
    since the properties of the \emph{unification} constraint merge
    can be proved independently from the \emph{subtyping} constraint merge.
\end{itemize}

\subsection{Least Upper Bound}
\label{sec:lub}

In this section, we present
the algorithm finding the least common supertype of two positive types. 
It is used directly by the constraint merge procedure (\cref{sec:constraint-merge}),
and indirectly, through the type upgrade by positive subtyping
(\cref{sec:subtyping}). Perhaps, the least upper bound is the least 
intuitive part of the algorithm, and its correctness will be covered
in \cref{sec:alg-upper-bounds-proofs}.

@@

\subsection{Anti-unification}
\label{sec:antiunification}

@@

\section{Declarative Typing}

In the previous section, we presented the 
type system together with subtyping specification 
and the algorithm. In this section, 
we describe the language under this type system, 
together with the type inference specification
and algorithm.

\subsection{Grammar}

First, we define the syntax of the language.
The language combines System F with call-by-push-value 
style. 
\begin{definition}[Language Grammar]
  \hfill
  \begin{multicols}{2}
    \ottgrammartabular{
      \ottc\ottinterrule
    }

    \ottgrammartabular{
      \ottv\ottinterrule
    }
  \end{multicols}
\end{definition}
Notice that the language does not have
first-class applications: instead, 
we use applicative let bindings---
constructions that bind a result of
a fully applied function to a (positive) 
variable.
In the call-by-push-value paradigm, 
it corresponds to monadic bind or do-notation. 
Typewise, these let-binders come in two forms:
annotated and unannotated. The annotated let-binders
$[[let x:iP = v(args); c]]$ 
requires the application to infer the annotated
$[[iP]]$, whereas the unannotated 
$[[let x = v(args); c]]$ 
is used when the inferred type is unique.

@@

\subsection{Declarative Type Inference}

@@

\begin{definition}[Declarative Type Inference]
  \label{def:declarative-typing}
  \ottdefnsDTLabeled
\end{definition}

@@

\section{Relation to System F}

@@

First, let us agree on the variant of \systemf that we use to establish the
relation with \fexists. At the type level, we have variables, functional arrows,
and universal quantifiers, and define existential quantifiers as a syntactic
sugar using standard encodings. At the term level, we have variables,
unannotated lambda abstractions, type abstractions, and term-level applications,
but not type applications (they can be done implicitly); for convenience, we
introduce multi-argument application, let-blinders, and existential constructors
and eliminators as syntactic sugar (see \cref{fig:systemf-grammar}).

@@

\subsection{Type-level Translation}

@@

\subsection{Term-level Translation}

@@

\paragraph{From \fexists to \systemf: Subtyping Elaboration}
The subtyping elaboration is defined in \cref{fig:subt-elaboration}.
The soundness property that is preserved by the elaboration is stated in
\cref{lemma:subt-elab-soundness}. Informally, $[[Γ ⊢ iN ≤ iM ⤳ ft]]$
guarantees that $[[ft]]$ is a \systemf term that represents 
a conversion (\ie a function) from the depolarized $[[|iN|]]$ to $[[|iM|]]$.
Symmetrically, $[[Γ ⊢ iP ≥ iQ ⤳ ft]]$ implies $[[|Γ| ; · ⊢ ft : |iQ| → |iP| ]]$.
\begin{itemize}
  \item Rules \ruleref{\ottdruleELNVarLabel} and \ruleref{\ottdruleELPVarLabel} 
    are trivial: if the left-hand side and the right-hand side types are the 
    same, the function converting one to another is the identity.
  \item Rules \ruleref{\ottdruleELShiftULabel} and \ruleref{\ottdruleELShiftDLabel}
    are symmetric. Since the shifts are removed during the depolarization,
    the conversion function is obtained by the recursive call to the 
    premise representing the subtyping of the required order
    (the mutual subtyping equivalence is defined as two separate judgments).
    Notice that the subtyping in the other order is still required
    so that the removal of the elaboration produces a correct
    subtyping inference tree in the original \fexists.
  \item Rule \ruleref{\ottdruleELArrowLabel} allows us to
    define elaboration between function. Since the subtyping is 
    contravariant on the argument type and covariant on the result type,
    we can use the elaboration functions acquired in the premises
    (their types are $[[|iQ| → |iP|]]$ and $[[|iN| → |iM|]]$)
    to construct the required elaboration term of type 
    $[[(|iP| → |iN|) → |iQ| → |iM|]]$.
  \item Rules 
    \ruleref{\ottdruleELForallLabel}, and \ruleref{\ottdruleELExistsLabel}
    are more involved. However, the soundness property is preserved
    by \cref{lemma:depol-subst} allowing one to distribute
    depolarization over the substitution whereby
    translating the instantiation substitution $[[σ]]$ from \fexists to \systemf.
\end{itemize}

\paragraph{From \fexists to \systemf: Typing Elaboration}
Using the subtyping elaboration, we define the elaboration for the
positive/negative typing and application typing in
\cref{fig:typing-elaboration}.
The soundness property (\cref{lemma:depolar-soundness})
guarantees that if the initial term $[[v]]$ has type $[[iP]]$ in \fexists,
then the result of the elaboration of this judgment $[[ft]]$ has
type $[[|iP|]]$ in \systemf. The negative typing has a symmetric property.
As with the subtyping elaboration, 
the rules in \cref{fig:typing-elaboration} are obtained from 
the \fexists declarative typing rules (\cref{def:declarative-typing})
by annotating each judgment with the elaboration term
so that the typing is preserved.

@@

\clearpage % This command will start a new page
\begin{figure}[t]
  \ottdefnTELNLabeledTwoColumns{}

  \hfill

  \ottdefnTELPAppLabeled{}
  \caption{Typing elaboration from \fexists to \systemf}
  \label{fig:typing-elaboration}
\end{figure}

\paragraph{From \systemf to \fexists}

@@

\section{Algorithmic Typing}

@@

\subsection{Algorithmic Type Inference}

Mirroring the declarative typing, 
the algorithm is represented by an inference system of three mutually recursive
judgments:
\begin{itemize}
  \item $[[Γ ; Φ ⊨ v : iP]]$ and  $[[Γ ; Φ ⊨ c : iN]]$
    are the algorithmic versions of $[[Γ ; Φ ⊢ v : iP]]$ and $[[Γ ; Φ ⊢ c : iN]]$.
    In contrast with the declarative counterparts, they are deterministic,
    and guarantee that the inferred type is normalized. 
  \item $[[Γ ; Φ ; Θ1 ⊨ uN ● args ⇒> uM ⫤ Θ2 ; SC]]$
    is the algorithmization of $[[Γ ; Φ ⊢ iN ● args ⇒> iM]]$.
    Notice that $[[uN]]$ contains algorithmic variables, 
    which are specified by the context $[[Θ1]]$.
    Moreover, the inferred type $[[uM]]$ is also algorithmic,
    and can have several non-equivalent instantiations. To accommodate that, 
    the algorithm also returns $[[Θ2]]$ and $[[SC]]$ specifying 
    the variables used in $[[uM]]$: $[[Θ2]]$ defines the contexts
    in which the variables must be instantiated, and $[[SC]]$
    imposes restrictions on the variables. 
\end{itemize}
As subroutines, the algorithm calls
subtyping (\cref{alg:subtyping}),
type well-formedness (\cref{alg:wf}),
constraint merge (\cref{sec:constraint-merge}),
normalization (\cref{alg:type-nf}),
and constraint singularity which will be defined later in 
\cref{sec:constraint-singularity}.
It also relies on basic set operations and the ability to 
deterministically choose fresh variables.

@@

The correctness of the algorithm consists of its soundness and 
completeness, which is proved by mutual
induction in \cref{lemma:typing-soundness,lemma:typing-completeness}.
The simplified result is the following.
\begin{theorempreview}
  \hfill
  \begin{itemize}
    \item [$-$] $[[Γ; Φ ⊨ c : iN]]$ implies $[[Γ; Φ ⊢ c : iN]]$, 
      and $[[Γ; Φ ⊢ c : iN]]$ implies $[[Γ; Φ ⊨ c : nf(iN)]]$;
    \item [$+$] $[[Γ; Φ ⊨ v : iP]]$ implies $[[Γ; Φ ⊢ v : iP]]$, 
      and $[[Γ; Φ ⊢ v : iP]]$ implies $[[Γ; Φ ⊨ v : nf(iP)]]$.
  \end{itemize}
\end{theorempreview}

\subsection{Minimal Instantiation}
The minimal instantiation algorithm is used
to infer the type of the bound variable in the un-annotated 
applicative let-binders, as long as there exists a principal
(minimal) type.  
Given a positive algorithmic type $[[uP]]$ 
and a set of constraints $[[SC]]$, it finds the substitution $[[uσ]]$
respecting $[[SC]]$ (i.e., $[[Θ ⊢ uσ : SC]]$) such that
it instantiates $[[uP]]$ to the minimal type, in other words 
for any other substitution $[[uσ']]$ respecting $[[SC]]$,
we have $[[Γ ⊢ [uσ']uP ≥ [uσ]uP]]$.

@@

\subsection{Constraint Singularity}
\label{sec:constraint-singularity}

@@

\section{Theorem Statements: Declarative}
\subsection{Type Well-Formedness}
\input{\proofGenDir/tex/wf-statements.tex}

\subsection{Substitution}
\input{\proofGenDir/tex/subst-statements.tex}

>>>>

% \renewcommand{\proofGenDir}{_gen}

\section{Definitions and Algorithms} 
\label{appendix}

\subsection{Declarative Types}

\subsubsection{Grammar}
We assume that there is an infinite set of positive and 
negative \emph{type} variables. Positive type variables are denoted as 
$[[α⁺]]$, $[[β⁺]]$, $[[γ⁺]]$, etc.
Negative type variables are denoted as $[[α⁻]]$, $[[β⁻]]$, $[[γ⁻]]$, etc.
We assume there is an infinite set of \emph{term} variables,
which are denoted as $[[x]]$, $[[y]]$, $[[z]]$, etc.
A list of objects (variables, types or terms) is denoted by
an overline arrow. For instance, $[[pas]]$ is a list of positive type variables, 
$[[nbs]]$ is a list of negative type variables, 
$[[args]]$ is a list of values, which are arguments of a function.
$[[fv(iP)]]$ and $[[fv(iN)]]$ denote the set of free variables 
in a type $[[iP]]$ and $[[iN]]$, respectively.
% \bigskip
% \ottgrammartabular{
%   \ottiP\ottinterrule
%   \ottiN\ottinterrule
% }

\begin{definition}[Declarative Types]
  \hfill\\
  \begin{minipage}[t]{0.45\textwidth}
    \ottgrammartabular{
      \ottiN\ottinterrule
    }
  \end{minipage}
  \hfill
  \begin{minipage}[t]{0.45\textwidth}
    \ottgrammartabular{
      \ottiP\ottinterrule
    }
  \end{minipage}
\end{definition}

\subsubsection{Equalities}
For simplicity, we assume alpha-equivalent terms are equal.
This way, we assume that substitutions do not capture bound variables.
Besides, we equate
$[[∀pas.∀pbs.iN]]$ with $[[∀pas,pbs.iN]]$, 
as well as $[[∃nas.∃nbs.iP]]$ with $[[∃nas,nbs.iP]]$,
and lift these equations transitively and congruently 
to the whole system.

\subsubsection{Contexts and Well-formedness}

@@

\subsubsection{Substitutions}

@@

\subsubsection{Declarative Subtyping}
Subtyping is one of the key mechanisms of our system. 
It realizes the polymorphism: abstract $[[∀]]$ and 
$[[∃]]$ types can be used where concrete types are expected,
exactly because of the subtyping relation between them.

@@

\subsection{Algorithmic Types}

\subsubsection{Grammar}

@@

\subsubsection{Fresh Variable Selection}
\label{sec:fresh-selection}
Both the subtyping and the type inference algorithm
rely on the ability to select fresh, unused variables.
For a set of variables $[[varset]]$, it is indicated as 
$[[varset are fresh]]$ in the inference rules.
We assume that the selection subroutine always succeeds and is 
deterministic. In other words, whenever it is called in 
an algorithmic inference rule, it returns the same result, 
uniquely determined by the input of this rule.

\subsubsection{Variable Algorithmization}
\label{sec:variable-algorithmization}

@@

\subsubsection{Contexts and Well-formedness}

@@

\subsubsection{Subsitutions}

@@

\subsubsection{Equivalence and Normalization}
\label{sec:equivalence-normalization}

@@

The rest of this chapter is devoted to the
central algorithm of the type system---the subtyping algorithm.

\subsubsection{Subtyping}
\label{sec:subtyping}

@@

Let us overview the inference rules of the subtyping algorithm.
\begin{itemize}
  \item \ruleref{\ottdruleANVarLabel} and \ruleref{\ottdruleAPVarLabel} 
    are the base cases. They copy the corresponding declarative rules and
    ensure reflexivity.
  \item \ruleref{\ottdruleAPUVarLabel} is the only case generating 
    subtyping constraints. In this case, we must ensure
    that the resulting constraints guarantee that the instantiation of 
    $[[â⁺]]$ is a supertype of $[[iP]]$.
    However, the obvious constraint $[[â⁺ :≥ iP]]$ might be problematic
    if  $[[iP]]$ is not well-formed in $[[Θ(â⁺)]]$. For this reason,
    we use the \emph{upgrade} procedure (it will be covered in \cref{sec:a:lub})
    to find the minimal supertype of $[[iP]]$, which is well-formed in $[[Θ(â⁺)]]$.

@@

  \item \ruleref{\ottdruleAShiftDLabel} and \ruleref{\ottdruleAShiftULabel} are the
    \emph{shift} rules. According to the declarative system,
    shifted subtyping requires equivalence. In the presence of the algorithmic 
    variables, it means that the left and the right-hand sides of the subtyping
    must be unified. Hence, the shift rules invoke the unification algorithm, 
    which will be discussed in \cref{sec:a:unification}. The unification 
    returns the minimal set of constraints $[[UC]]$, which is necessary
    and sufficient for the subtyping.

  \item \ruleref{\ottdruleAArrowLabel}.
    In this case, the algorithm makes two calls:
    a recursive call to the negative subtyping algorithm for the argument types,
    and a call to the positive subtyping algorithm for the result types.
    After that, the resulting constraints are merged using the
    \emph{subtyping constraint merge} procedure, 
    which is discussed in \cref{sec:a:constraint-merge}.
  \item \ruleref{\ottdruleAForallLabel} and \ruleref{\ottdruleAExistsLabel}
    are symmetric. These are the only places where 
    the algorithmic variables are introduced.
    It is done by algorithmization (\cref{sec:variable-algorithmization}) 
    of the quantified variables: these variables are replaced by 
    fresh algorithmic variables in the body of the quantified type,
    the algorithmic variables are added to the subtyping context $[[Θ]]$,
    after that, the recursive call is made. Notice that the declarative context
    $[[Γ]]$ is extended by the quantified variables from the right-hand side,
    which matches the declarative system.
\end{itemize}

@@

\subsubsection{Constraints}
\label{sec:a:constraint-merge}

@@

\paragraph{Constraint Satisfaction}

@@

\paragraph{Constraint Merge}

@@

  \item \ruleref{\ottdruleSCMESupSupLabel} 
    In this case, we find the least upper bound of the types from the input
    restrictions, 
    and as the output, restrict the variable to be a supertype of the result.
    The least upper bound procedure will be discussed in \cref{sec:a:lub}.
\end{itemize}

@@

\subsubsection{Unification}
\label{sec:a:unification}

@@

\begin{itemize}
  \item \ruleref{\ottdruleUShiftULabel}, \ruleref{\ottdruleUShiftDLabel}, 
    \ruleref{\ottdruleUForallLabel}, and \ruleref{\ottdruleUExistsLabel}
    are defined congruently. In the shift rules, the algorithm
    removes the outermost constructor. In the
    $[[∀]]$ and $[[∃]]$ rules, it removes the quantifiers,
    adding the quantified variables to the context $[[Γ]]$.
    Notice that $[[Θ]]$, which specifies
    the contexts in which the algorithmic variables must be instantiated,
    is not changed.
  \item \ruleref{\ottdruleUNVarLabel} and \ruleref{\ottdruleUPVarLabel} 
    are the base cases. 
    Since the sides are equal and free from algorithmic variables,
    the unification returns an empty constraint. 
  \item \ruleref{\ottdruleUNVarLabel} and \ruleref{\ottdruleUPVarLabel}
    are symmetric cases constructing the constraints. 
    When an algorithmic variable is unified with a type, 
    we must check that the type is well-formed in the required context,
    and if it is, we return a constraint restricting the variable
    to be equivalent to that type.
  \item \ruleref{\ottdruleUArrowLabel}.
    In this case, the algorithm makes two recursive calls:
    it unifies the arguments and the results of the arrows.
    After that, the resulting constraints are merged using the
    \emph{unification constraint merge} procedure, 
    which is discussed in \cref{sec:a:constraint-merge}.
    Notice that $[[UC1]]$ and $[[UC2]]$ are guaranteed to be
    \emph{unification} constraints, not arbitrary \emph{subtyping} 
    constraints: it is important for modularizing the proofs, 
    since the properties of the \emph{unification} constraint merge
    can be proved independently from the \emph{subtyping} constraint merge.
\end{itemize}

\subsubsection{Least Upper Bound}
\label{sec:a:lub}

In this section, we present
the algorithm finding the least common supertype of two positive types. 
It is used directly by the constraint merge procedure (\cref{sec:a:constraint-merge}),
and indirectly, through the type upgrade by positive subtyping
(\cref{sec:subtyping}). Perhaps, the least upper bound is the least 
intuitive part of the algorithm, and its correctness will be covered
in \cref{sec:alg-upper-bounds-proofs}.

@@

\subsubsection{Anti-unification}
\label{sec:a:antiunification}

@@

\subsection{Declarative Typing}

In the previous section, we presented the type system together with the subtyping
specification and the algorithm. In this section, we describe the language under
this type system, together with the type inference specification and algorithm.

\subsubsection{Grammar}

The syntax of \fexists terms is given by the following grammar:
\begin{definition}[Grammar of Terms]
  \hfill\\
  \begin{minipage}[t]{0.45\textwidth}
    \ottgrammartabular{
      \ottc\ottinterrule
    }
  \end{minipage} \hfill
  \begin{minipage}[t]{0.45\textwidth}
    \ottgrammartabular{
      \ottv\ottinterrule
    }
  \end{minipage}
\end{definition}

Notice that the language does not have
first-class applications: instead, 
we use applicative let bindings---
constructions that bind a result of
a fully applied function to a (positive) 
variable.
In the call-by-push-value paradigm, 
it corresponds to monadic bind or do-notation. 
Typewise, these let-binders come in two forms:
annotated and unannotated. The annotated let-binders
$[[let x:iP = v(args); c]]$ 
requires the application to infer the annotated
$[[iP]]$, whereas the unannotated 
$[[let x = v(args); c]]$ 
is used when the inferred type is unique.

@@

\subsubsection{Declarative Type Inference}

@@

\begin{definition}[Declarative Type Inference]
  \hfill\\
  \label{def:declarative-typing}
  \ottdefnsDTLabeled
\end{definition}

@@

\subsection{Relation between \fexists and \systemf}

@@

First, let us agree on the variant of \systemf that we use to establish the
relation with \fexists. At the type level, we have variables, functional arrows,
and universal quantifiers, and define existential quantifiers as a syntactic
sugar using standard encodings. At the term level, we have variables,
unannotated lambda abstractions, type abstractions, and term-level applications,
but not type applications (they can be done implicitly); for convenience, we
introduce multi-argument applications, let-blinders, and existential constructors
and eliminators as syntactic sugar (see \cref{fig:systemf-grammar}).

@@

\subsubsection{Type-level Translation}

@@

\subsubsection{Term-level Translation}

\label{sec:term-level-translation}

@@

The subtyping elaboration is defined in \cref{fig:subt-elaboration}.
The soundness property that is preserved by the elaboration is stated in
\cref{lemma:subt-elab-soundness}. Informally, $[[Γ ⊢ iN ≤ iM ⤳ ft]]$
guarantees that $[[ft]]$ is a \systemf term that represents 
a conversion (\ie a function) from the depolarized $[[|iN|]]$ to $[[|iM|]]$.
Symmetrically, $[[Γ ⊢ iP ≥ iQ ⤳ ft]]$ implies $[[|Γ| ; · ⊢ ft : |iQ| → |iP| ]]$.
\begin{itemize}
  \item Rules \ruleref{\ottdruleELNVarLabel} and \ruleref{\ottdruleELPVarLabel} 
    are trivial: if the left-hand side and the right-hand side types are the 
    same, the function converting one to another is the identity.
  \item Rules \ruleref{\ottdruleELShiftULabel} and \ruleref{\ottdruleELShiftDLabel}
    are symmetric. Since the shifts are removed during the depolarization,
    the conversion function is obtained by the recursive call to the 
    premise representing the subtyping of the required order
    (the mutual subtyping equivalence is defined as two separate judgments).
    Notice that the subtyping in the other order is still required
    so that the removal of the elaboration produces a correct
    subtyping inference tree in the original \fexists.
  \item Rule \ruleref{\ottdruleELArrowLabel} allows us to
    define elaboration between function. Since the subtyping is 
    contravariant on the argument type and covariant on the result type,
    we can use the elaboration functions acquired in the premises
    (their types are $[[|iQ| → |iP|]]$ and $[[|iN| → |iM|]]$)
    to construct the required elaboration term of type 
    $[[(|iP| → |iN|) → |iQ| → |iM|]]$.
  \item Rules 
    \ruleref{\ottdruleELForallLabel}, and \ruleref{\ottdruleELExistsLabel}
    are more involved. However, the soundness property is preserved
    by \cref{lemma:depol-subst} allowing one to distribute
    depolarization over the substitution whereby
    translating the instantiation substitution $[[σ]]$ from \fexists to \systemf.
\end{itemize}

Using the subtyping elaboration, we define the elaboration for the
positive/negative typing and application typing in
\cref{fig:typing-elaboration}.
The soundness property (\cref{lemma:depolar-soundness})
guarantees that if the initial term $[[v]]$ has type $[[iP]]$ in \fexists,
then the result of the elaboration of this judgment $[[ft]]$ has
type $[[|iP|]]$ in \systemf. The negative typing has a symmetric property.
As with the subtyping elaboration, 
the rules in \cref{fig:typing-elaboration} are obtained from 
the \fexists declarative typing rules (\cref{def:declarative-typing})
by annotating each judgment with the elaboration term
so that the typing is preserved.

@@

\clearpage
\begin{figure}[t]
  \ottdefnTELNLabeledTwoColumns{}

  \ottdefnTELPAppLabeled{}
  \caption{Typing elaboration from \fexists to \systemf}
  \label{fig:typing-elaboration}
\end{figure}
\clearpage

@@

\subsection{Algorithmic Typing}

@@

\subsubsection{Algorithmic Type Inference}

Mirroring the declarative typing, 
the algorithm is represented by an inference system of three mutually recursive
judgments:
\begin{itemize}
  \item $[[Γ ; Φ ⊨ v : iP]]$ and  $[[Γ ; Φ ⊨ c : iN]]$
    are the algorithmic versions of $[[Γ ; Φ ⊢ v : iP]]$ and $[[Γ ; Φ ⊢ c : iN]]$.
    In contrast with the declarative counterparts, they are deterministic,
    and guarantee that the inferred type is normalized. 
  \item $[[Γ ; Φ ; Θ1 ⊨ uN ● args ⇒> uM ⫤ Θ2 ; SC]]$
    is the algorithmization of $[[Γ ; Φ ⊢ iN ● args ⇒> iM]]$.
    Notice that $[[uN]]$ contains algorithmic variables, 
    which are specified by the context $[[Θ1]]$.
    Moreover, the inferred type $[[uM]]$ is also algorithmic,
    and can have several non-equivalent instantiations. To accommodate that, 
    the algorithm also returns $[[Θ2]]$ and $[[SC]]$ specifying 
    the variables used in $[[uM]]$: $[[Θ2]]$ defines the contexts
    in which the variables must be instantiated, and $[[SC]]$
    imposes restrictions on the variables. 
\end{itemize}
As subroutines, the algorithm calls
subtyping (\cref{alg:subtyping}),
type well-formedness (\cref{alg:wf}),
constraint merge (\cref{sec:a:constraint-merge}),
normalization (\cref{alg:type-nf}),
and constraint singularity which will be defined later in 
\cref{sec:a:constraint-singularity}.
It also relies on basic set operations and the ability to 
deterministically choose fresh variables.

@@

The correctness of the algorithm consists of its soundness and 
completeness, which is proved by mutual
induction in \cref{lemma:typing-soundness,lemma:typing-completeness}.
The simplified result is the following.
\begin{theorempreview}[Correctness of Algorithmic Typing (simplified)]
  \hfill
  \begin{itemize}
    \item [$-$] $[[Γ; Φ ⊨ c : iN]]$ implies $[[Γ; Φ ⊢ c : iN]]$, 
      and $[[Γ; Φ ⊢ c : iN]]$ implies $[[Γ; Φ ⊨ c : nf(iN)]]$;
    \item [$+$] $[[Γ; Φ ⊨ v : iP]]$ implies $[[Γ; Φ ⊢ v : iP]]$, 
      and $[[Γ; Φ ⊢ v : iP]]$ implies $[[Γ; Φ ⊨ v : nf(iP)]]$.
  \end{itemize}
\end{theorempreview}

\subsubsection{Minimal Instantiation}
The minimal instantiation algorithm is used
to infer the type of the bound variable in the un-annotated 
applicative let-binders, as long as there exists a principal
(minimal) type.  
Given a positive algorithmic type $[[uP]]$ 
and a set of constraints $[[SC]]$, it finds the substitution $[[uσ]]$
respecting $[[SC]]$ (i.e., $[[Θ ⊢ uσ : SC]]$) such that
it instantiates $[[uP]]$ to the minimal type, in other words 
for any other substitution $[[uσ']]$ respecting $[[SC]]$,
we have $[[Γ ⊢ [uσ']uP ≥ [uσ]uP]]$.

@@

\subsubsection{Constraint Singularity}
\label{sec:a:constraint-singularity}

@@

\clearpage
\section{Theorem Statements}

\subsection{Theorem Statements: Declarative}
\subsubsection{Type Well-Formedness}
\input{\proofGenDir/tex/wf-statements.tex}

\subsubsection{Substitution}
\input{\proofGenDir/tex/subst-statements.tex}
